Pre-Main-Sequence M-Dwarfs With Complex Quasiperiodic Variability From Four Years of TESS

### 2023-06-16 paper/ms.tex
update ms

% 250 word max!
\begin{abstract}
  Complex quasiperiodic variables (CQVs) are pre-main-sequence
  M-dwarfs with almost-periodic optical modulation that includes both
  sharp and broad troughs.  The troughs are usually superposed over
  smooth, spot-like modulation.  These objects are typically
  10-100 million years (Myr) old, and they do not seem to have primordial gas-rich
  disks.  Nonetheless, the preferred explanations for their optical
  variability invoke long-lived material in orbit at the corotation
  radius, for instance transiting dust clumps or stellar prominences.
  Here, we report new CQVs discovered in TESS data collected between
  July~2018 and Sep~2022.  Our search used two methods, the first
  based on counting dips in phase-folded light curves, and the second
  based on counting peaks in the Fourier domain.  From our analysis of
  \nstarssearched\ M-dwarfs with $T$$<$16 and $d$$<$150\,pc, we find
  \ngoods\ gold-standard CQVs.  Most of these discoveries are new,
  and they include the brightest and closest
  examples of this class of object known ($T$$\approx$10; $d$$\approx$20\,pc).  A few
  objects are outliers among outliers; LP 12-502 for instance
  shows four to eight local minima per cycle distributed over a ``dip
  complex'' with a total duration that is nearly fixed over more than
  1000 cycles.  Nonetheless, LP-502 at times also showed drastic shape
  changes over less than one cycle.  More generally, we find that none
  of the CQVs maintain a fixed light curve shape over timescales of
  more than a few hundred cycles -- they are {\it quasi}periodic.
  Although the physical explanation behind these objects remains
  enigmatic, our sample may help clarify their origin.  Connections to
  star, disk, and exoplanet evolution are emphasized.
\end{abstract}

@@

The issue of what exactly causes the variability that defines the CQVs
has yet to be resolved.  There are a few competing explanations.  All
young M-dwarfs are spotted, which produces flux variations over
characteristic timescales of the rotation period, $P_{\rm rot}$, and
its half-harmonic, $0.5\,P_{\rm rot}$.  The observed dips occur over
durations as short as $0.05\,P_{\rm rot}$; a ``starspot-only''
scenario is therefore highly unlikely for any of the objects with
sufficiently sharp dips
\citep{2017AJ....153..152S,2021MNRAS.500.1366K}.   The more likely
scenarios instead invoke sharp geometries with material extrinsic to
the stellar surface \citep[e.g.][]{2017AJ....153..152S,2022AJ....163..144G}.  In the ``clump''
scenario, opaque dust clumps orbiting near the Keplerian corotation
radius eclipse the star
\citep{2017AJ....153..152S,2023MNRAS.518.4734S}.  In the
``prominence'' scenario, long-lived condensations of plasma are
trapped along the star’s magnetic field lines, also near corotation
\citep{2022MNRAS.514.5465W}.  And in a ``screen'' scenario, the inner
wall of a quiescent circumstellar disk blocks a portion of the stellar
surface to produce sudden dips whenever spots come into view
\citep{2019ApJ...876..127Z}.  In our view, the arguments seem best
established for the dust clump and prominence hypotheses
\citep{2023MNRAS.518.4734S,2022MNRAS.514.5465W}.  However, unambiguous
evidence has yet to be presented to demonstrate which, if any, of
these scenarios is correct.

@@

\subsubsection{Fourier analysis}
\label{subsec:fourier}
For the Fourier analysis, we followed
\citet{2019ApJ...876..127Z}.
{\bf todo for Rahul or Saul: explain the approach, in a few
paragraphs.  Was the SAP\_FLUX or PDCSAP used? etc. }

@@

\paragraph{Ages}
We estimated the stellar ages by making probabilistic spatial and
kinematic associations between the CQVs and known clusters in the solar
neighborhood.  For most of the stars in our sample, we were able to do
this using BANYAN\,$\Sigma$
\citep{2018ApJ...856...23G}.\footnote{\url{https://github.com/jgagneastro/banyan_sigma},
git commit \texttt{394b486}} This algorithm calculates the
probability that a given star belongs to one of 27 young
clusters (or ``assocations'') within 150\,pc of the Sun, by modeling the clusters
as multivariate Gaussians in 3-D position and 3-D velocity space.  
We used the
Gaia DR2 sky positions, proper motions, and distances to calculate the
membership probabilities.  
BANYAN\,$\Sigma$ in turn analytically marginalizes over the radial velocity dimension.
We emphasize that the probabilities returned by this procedure are qualitatively
useful, but quantitatively dubious due to the non-Gaussian
nature of most groups within the
solar neighborhood \citep[see e.g.][Figure~10]{2021ApJ...917...23K}.

@@

Table~\ref{tab:thetable} lists the \ncpvsfound\ CQVs identified by
our search, and some of their important properties.
\ngoods\ objects demonstrate what we view as unambiguous
characteristics of the CPV phenomenon in at least one TESS sector, and
we refer to these as the ``gold sample''.
The remaining \nmaybes\ CPV candidates were ambiguous (in our view),
and their variability might be caused by say, multiple starspot
groups.

@@

Of our \ncpvsfound\ confirmed and candidate CQVs, \nnotfieldbanyan\ 
were confidently associated with a nearby moving group or open
cluster using BANYAN\,$\Sigma$.
The relevant groups are listed in Table~\ref{tab:thetable}; their ages
span $\approx$5-150\,Myr.
The most prodigious groups were Sco-Cen, Tuc-Hor, and Columba.
The yield of CQVs in Sco-Cen (LCC / UCL / USco) is not surprising,
since Sco-Cen contains the majority of pre-main-sequence stars in the
solar neighborhood.
The absence of CQVs in IC2602 and IC2391 is because those clusters are
beyond 100\,pc, where the completeness of the TESS 120-second cadence
data is below 10\%.

Of the the \ngoodsfieldbanyan\ non-associated stars, one
(TIC~302160226) is an obvious member of $\alpha$~Per ($t\approx
86\pm16$\,Myr; \citealt{2022arXiv221109822B}).  For the other two
(TIC~58084670 and TIC~141146667), we are convinced of both their CPV
status and their youth, but we were not able to confidently associate
either star with any young groups.
Both were noted as being in the ``diffuse'' Class~II YSO
population by \citet{2021ApJ...917...23K}.
That diffuse population may contain kinematic structures that have yet
to be unravelled, however the absence of obvious young, co-moving
stars within $\Delta r$=$40$\,pc and $\Delta v$=$3$\,km\,s$^{-1}$
of these stars suggest that they may have run away from their birth
locations.

@@

\paragraph{LP 12-502}
While many CPVs had multiple sectors of data, in our view the most
astounding dataset was available for LP 12-502 (TIP 402980664).
Figure~\ref{fig:lp} shows all available data from Sectors 18, 19, 25,
26, 53, 58, and 59; the star was observed at 120-second cadence for
every sector during which it was on silicon.

For the first 64 cycles, the star shows a pattern reminiscent of a
(bizarre) eclipsing binary, with four obvious dips, that we denote
dips 1, 2, 3, and 4 at phases -0.28, -0.08, 0, and 0.25 respectively.
The depth of dips 1 and 3 remain roughly fixed.
Over the 64 cycles, dip 4 {\it decreases} in depth by about 2\%, while
dip 2 {\it increases} in depth, by about the same amount.

There is then a 6-month (184 cycle) gap out to Cycle 248.
At this time, dip 1 is still present, at the same phase, and with
roughly the same depth.

%TODO continue writing about this.  What is the exact TJD of the jump?

@@

\begin{figure*}[!t]
	\begin{center}
    \centering
    \includegraphics[width=0.9\textwidth]{TIC_402980664_P18.5604_2min_phase_timegroups_mosaic.pdf}
		\vspace{-0.45cm}
		\caption{
			{\it Bottom:}
      Evolution of LP 12-502 ($P$=18.5604\,h) at fixed period and epoch over three
      years; small text denotes cycle number.  The TESS pointing law
      dictates time gaps; larger gaps tend to yield larger shape
      changes.
      The dips usually evolve over tens to hundreds of cycles.
      However cycles 1233-1264 show a dip that ``switched'' from a depth
      and duration of 3\% and 3\,hr to 0.3\% and 1\,hr over less than
      one cycle.
		}
		\label{fig:lp}
	\end{center}
\end{figure*}

@@

{\bf author contribution statement goes here}
L.G.B.~and R.~J. conceived the project and executed the
dip-based and Fourier-based searches, respectively.
They both drafted the initial manuscript.
L.G.B.~performed the cluster membership analysis.
S.~R. vetted the results from the Fourier search.
L.A.H.~contributed to project design.
All authors assisted in manuscript revision.

>>>>

% 250 word max!
\begin{abstract}
  Complex quasiperiodic variables (CQVs) are pre-main-sequence
  M-dwarfs with almost-periodic optical modulation that includes both
  sharp and broad troughs.  The troughs are usually superposed over
  smooth, spot-like modulation.  These objects are typically 10-100
  million years (Myr) old, and they do not have primordial gas-rich
  disks.  Nonetheless, the preferred explanations for their optical
  variability invoke long-lived material in orbit at the Keplerian
  corotation radius, for instance transiting dust clumps or stellar
  prominences.  Here, we report new CQVs discovered in TESS data
  collected between July~2018 and Sep~2022.  Our search used two
  methods, the first based on counting dips in phase-folded light
  curves, and the second based on counting peaks in the Fourier
  domain.  From our analysis of \nstarssearched\ M-dwarfs with
  $T$$<$16 and $d$$<$150\,pc, we find \ngoods\ gold-standard CQVs.
  Most of these discoveries are new, and they include the brightest
  and closest examples of this class of object known ($T$$\approx$11;
  $d$$\approx$20\,pc).  A few objects are outliers among outliers; LP
  12-502 for instance shows four to eight local minima per cycle
  distributed over a ``dip complex'' with a total duration that is
  nearly constant over more than 1000 cycles.  Nonetheless, LP-502 at
  times also showed drastic shape changes over less than one cycle.
  More generally, we find that none of the CQVs maintain a fixed light
  curve shape over timescales of more than a few hundred cycles --
  they are {\it quasi}periodic.  Although the physical explanation
  behind these objects remains enigmatic, our sample may help clarify
  their origin.  Connections to star, disk, and exoplanet evolution
  are emphasized.
\end{abstract}

@@

The issue of what causes the variability that defines the CQVs has yet
to be resolved.  There are a few competing explanations.  All young
M-dwarfs are spotted, which produces flux variations over
characteristic timescales of the rotation period, $P_{\rm rot}$, and
its half-harmonic, $0.5\,P_{\rm rot}$.  The observed dips occur over
durations as short as $0.05\,P_{\rm rot}$; a ``starspot-only''
scenario is therefore highly unlikely for any of the objects with
sufficiently sharp dips
\citep{2017AJ....153..152S,2021MNRAS.500.1366K}.   The more likely
scenarios instead invoke sharp geometries with material extrinsic to
the stellar surface
\citep[e.g.][]{2017AJ....153..152S,2022AJ....163..144G}.  In the
``clump'' scenario, opaque dust clumps orbiting near the Keplerian
corotation radius eclipse the star
\citep{2017AJ....153..152S,2023MNRAS.518.4734S}.  In the
``prominence'' scenario, long-lived condensations of plasma are
trapped along the star’s magnetic field lines, also near corotation
\citep{2022MNRAS.514.5465W}.  And in a ``screen'' scenario, the inner
wall of a quiescent circumstellar disk blocks a portion of the stellar
surface to produce sudden dips whenever spots come into view
\citep{2019ApJ...876..127Z}.  In our view, the arguments seem best
established for the dust clump and prominence hypotheses
\citep{2023MNRAS.518.4734S,2022MNRAS.514.5465W}.  However, unambiguous
evidence has yet to be presented to demonstrate which, if any, of
these scenarios is correct.

@@

\subsubsection{Fourier analysis}
\label{subsec:fourier}
For the Fourier analysis, we followed \citet{2019ApJ...876..127Z}.

{\bf todo for Rahul or Saul: explain the approach, in a few
paragraphs.  Was the SAP\_FLUX or PDCSAP used? etc. }

@@

\paragraph{Ages}
We estimated the stellar ages by making probabilistic spatial and
kinematic associations between the CQVs and known clusters in the
solar neighborhood.  For most of the stars in our sample, we were able
to do this using BANYAN\,$\Sigma$
\citep{2018ApJ...856...23G}.\footnote{\url{https://github.com/jgagneastro/banyan_sigma},
git commit \texttt{394b486}} This algorithm calculates the probability
that a given star belongs to one of 27 young clusters (or
``assocations'') within 150\,pc of the Sun, by modeling the clusters
as multivariate Gaussians in 3-D position and 3-D velocity space.  We
used the Gaia DR2 sky positions, proper motions, and distances to
calculate the membership probabilities.  BANYAN\,$\Sigma$ in turn
analytically marginalizes over the radial velocity dimension.  The
probabilities returned by this procedure are qualitatively useful, but
quantitatively dubious due to the non-Gaussian nature of most groups
within the solar neighborhood \citep[see
e.g.][Figure~10]{2021ApJ...917...23K}.

@@

Table~\ref{tab:thetable} lists the \ncpvsfound\ CQVs identified by our
search, and some of their important properties.  \ngoods\ objects
demonstrate what we view as unambiguous characteristics of the CPV
phenomenon in at least one TESS sector, and we refer to these as the
``gold sample''.  The remaining \nmaybes\ CPV candidates were
ambiguous (in our view), and their variability might be caused by say,
multiple starspot groups.  Additional TESS data could help resolve
their classification.

@@

Of our \ncpvsfound\ confirmed and candidate CQVs, \nnotfieldbanyan\
were confidently associated with a nearby moving group or open cluster
using BANYAN\,$\Sigma$.  The relevant groups are listed in
Table~\ref{tab:thetable}; their ages span $\approx$5-150\,Myr.  The
most prodigious groups were Sco-Cen, Tuc-Hor, and Columba.  The yield
in Sco-Cen (LCC / UCL / USco) is not surprising, since Sco-Cen
contains the majority of pre-main-sequence stars in the solar
neighborhood.  The absence of CQVs in IC\,2602 and IC\,2391 can be
understand by those clusters being beyond 100\,pc, where the
completeness of the TESS 120-second cadence data is below 10\%; based
on the abundance of CQVs in the 40\,Myr Tuc-Hor and Columba, the
similarly-aged IC\,2602 and IC\,2391 clusters should have a large
number of CQVs in the TESS full frame image data.

Of the \ngoodsfieldbanyan\ stars for which BANYAN\,$\Sigma$ did not
find any association, one (TIC~302160226) is a member of $\alpha$~Per
($t\approx 86\pm16$\,Myr;
\citealt{2021A&A...645A..84M,2022arXiv221109822B}).  For the other two
(TIC~58084670 and TIC~141146667), we are convinced of both their CPV
status and their youth, but we were not able to confidently associate
either star with any young groups.  Both objects were noted as being in the
``diffuse'' Class~II YSO population near the Sun by \citet{2021ApJ...917...23K}.
That diffuse population may contain kinematic structures that have yet
to be unravelled.  However, the absence of obvious young, co-moving
stars within $\Delta r$=$40$\,pc and $\Delta v$=$3$\,km\,s$^{-1}$ of
our two CPVs suggest that they may have ``run away'' from their birth
locations.

@@

\paragraph{LP 12-502}
While many CPVs had multiple sectors of data, an astounding dataset
was available for LP 12-502 (TIC 402980664; $d$=21\,pc, $J$=9.4,
$T$=11.1).  Figure~\ref{fig:lp} shows all available data from Sectors
18, 19, 25, 26, 53, 58, and 59; the star was observed at 120-second
cadence for every sector during which it was on silicon.

For the first 64 cycles, the star shows a pattern reminiscent of a
(bizarre) eclipsing binary, with four obvious dips, that we dub
dips $\{ 1, 2, 3, 4 \}$ at phases $\{ -0.28, -0.08, 0, 0.25 \}$
respectively.  Over cycles 0 to 64, the depth of dips 1 and 3 remain
roughly fixed.  However dip 4 {\it decreases} in depth by about 2\%,
while dip 2 {\it increases} in depth, by about the same amount.

{\bf todo: discussion needs auto-measured phases at fixed t0...}
There is then a 6-month (184 cycle) gap out to Cycles 248-315, which
show an intricately structured dip complex.
Between Cycles 248-264, there are at least six dips, at phases $\{
  -0.47, -0.28, -0.08,  0.  ,  0.06,  0.17 $\}.
There is ``state switch'', roughly at BTJD 1993.5 - cycle 262.
The lower black points in cycles 248--264 shift upward during cycle
262, never to return.
A large flare occurred a few cycles earlier, at BTJD 1989.2 (which may
or may not be related).
Regardless, the depths of the latter three-dip complex remain roughly
the same after the state-switch;
the first obvious dip in the sequence disappears.
Ah, and the flux changes by 3.5\% over about 22 minutes (0.02\,$P$),
which is the steepest slope exhibited anywhere in this object's
remarkable dataset.
During the beginning of Sector~26 (Cycles 283-298), there are seven
clear dips per cycle.
The the second orbit of this sector,
there is another state-switch at BTJD 2030.7, immediately during a
flare.   (And therefore likely caused by the flare!)

Finally, sectors 53--58 (cycles 1233-1481) showed between four and six
dips per cycle, at various times.  
Some dips, for instance the one at $\phi = -0.03$, remain stable in
depth and duration over this five month interval.
Other dips, such as the one at $\phi = +0.04$ in cycles 1233-1264,
disappear entirely.
Other dips grow, like the one at $\phi = +0.06$ between cycles 1499
and 1481.
Also, there is the epic state switch during cycles 1233-1247.
Which appears to happen over less than one cycle, and there is no
associated flare.

Some are sharktooths.
Dip1 in cycle48-64 is one.
Dip1 in cycle 1466-1481 is one.
It is definitely asymmetric!

@@

\begin{figure*}[!t]
	\begin{center}
    \centering
    \includegraphics[width=0.9\textwidth]{TIC_402980664_P18.5604_2min_phase_timegroups_mosaic.pdf}
		\vspace{-0.45cm}
		\caption{
			{\it Bottom:}
      Evolution of LP 12-502 ($P$=18.5604\,h) at fixed period and
      epoch over three years; small text denotes cycle number.  
      There are 200 binned black points per cycle.
      The TESS pointing law dictates time gaps; larger gaps tend to
      yield larger shape changes.
      The dips usually evolve over tens to hundreds of cycles.
      However cycles 1233-1264 show a dip that ``switched'' from a depth
      and duration of 3\% and 3\,hr to 0.3\% and 1\,hr over less than
      one cycle.
		}
		\label{fig:lp}
	\end{center}
\end{figure*}

@@

{\bf author contribution statement goes here}
L.G.B.~and R.~J. conceived the project and executed the
dip-based and Fourier-based searches, respectively.
L.G.B.~drafted the initial manuscript, and performed the cluster
membership analysis.
S.~R. and R.~J. vetted the results from the Fourier search.
L.A.H.~contributed to project design.
All authors assisted in manuscript revision.
